# Power calculation


::: {.callout-tip}
## Learning objectives

The learning objectives for this theme is to understand the idea behind statistical power.

* Know that a true difference might not be statistical detected due to size of data. 
* Understand that statistical power depends on effect size, uncertainty and number of samples. 
* Be able to make appropriate assumptions and calculate the power for a given study design. 
* Be able to calculate power for studies evaluated by the t-test and by the binomial distribution. 
:::

::: {.callout-example #exa-qa-power-calculation}
### Quality control - power calculation

This example extends @exa-qa-estimation.

The $3\%$ nonconforming products is unsatisfactory, and so you use a lot of money and (hopefully) make a lot of improvements. After a year you and your colleagues feel that the end quality has improved, and you wish to test that this improvement is indeed also statistically provable.

You believe that the number of nonconforming products is reduced by a factor of $2$, that is down to $1.5\%$ nonconforming products.

The question is: In a trial, how many samples ($n$) should you select in order to statistically prove this change?.

The null hypothesis is $H0: p = p_0 = 0.03$.

With the *one-sided* alternative: $HA: p < p_0 = 0.03$.

From a study on say $n = 200$ ($X_{HO} \sim \mathcal{B}(n,p = 0.03)$) with $x$ nonconforming products the probability for accepting $H0$ is: 

\begin{equation}
P(X_{HO} \leq x) \geq \alpha = 0.05
\end{equation}

```{r}

n <- 200
x <- 0:3
pbinom(x, n, 0.03)

```

So, in order to *reject* $H0$ (at level $\alpha = 0.05$) you should among $200$ samples find $0$ or $1$ nonconforming products (at $x = 2$, the p-value is $p = 0.059$).

If you believe that the true probability is $p = 0.015$ ($X_{HA} \sim \mathcal{B}(n,p = 0.015)$) then the probability for getting this number of positive samples (or what is even more extreme compared to $H0$) can be calculated as

\begin{equation}
P(X_{HA} \leq 1) = P(X_{HA}=0) + P(X_{HA}=1).
\end{equation}

```{r}
n <- 200
x <- 1
pbinom(x, n, 0.015)
```

**This is the power of the trial** under these assumptions.
In detail that means, that only one out of five times you would be able to prove that the investment were worth the effort.

If you which to have a higher power for the study, then the number of samples should be increased e.g. to $n = 400$:

```{r}
alpha <- 0.05
n <- 400 # Number of trials
x <- 0:100 # Number of successes

# The probability of outcome x under the null hypothesis that p = 0.03
nullProb <- pbinom(x, n, 0.03) 


# The maximum x-value that returns a probability smaller than alpha
xMax <- max(x[nullProb < alpha])

pbinom(xMax, n, 0.015)
```

Or assessed for a sequence of $n$ values (@fig-power-vs-n).
```{r}
#| label: fig-power-vs-n
#| fig-cap: Power vs number of trials.
#| warning: false


alpha <- 0.05
N <- 1:800 # Number of trials
pwr <- c() # Initiate vector for storage


for (n in N) {
  
  x <- 0:n
  
  nullProb <- pbinom(x, n, 0.03)
  
  xMax <- max(x[nullProb < alpha])
  pwr[n] <- pbinom(xMax, n, 0.015)
}

plot(N, pwr)

```

It is seen that indeed a very high number of samples needs to be collected in order to be fairly certain that the trial will statistically confirm that the number of nonconforming products has dropped.


:::
<!-- End example callout -->