## Rmarkdown

Rmarkdown is a framework for producing beautiful reports where code and text is combined side-by-side. It makes it possible to combine scripts (your data analysis) with output (figures, tables and numbers) and narrative (the fairytale on why and discussions etc.) in ONE document. It is rather simple when you get used to how it works, and it really makes life much more easy. Both for this course, but also for all other tasks which involves data analysis and is to be presented as a report.

It is refereed to as reproducible data analysis, and is the opposite of Excel-Hell, where the latter is characterized by being non transparent, and really hard to figure out what happened between data and results.

::: {.callout-tip}
    Rmarkdown is especially recommended for the hand-in assignemnts doing the course. It is much easier than copying all plots, outputs etc. to a Word document.
:::

### Setting up a Rmarkdown document

To setup a new Rmarkdown document do the following:

1. In RStudio go to File -> New File -> RMarkdown
2. In the popup set your title, author name and date (these can all be changed later).
3. Choose your default output. HTML or PDF is recommended (this can be changed later).
4. Click "OK".

Now your new document has been created. It contains some settings at the very top:

```{r}
#| eval: false
---
title: "My first project"
author: "Jeppe"
date: "`r Sys.Date()`"
output: html_document # Change this to pdf_document for pdf output
---

```

These settings can be changed whenever. For more info on fancy settings see [guide on HTML outputs](https://bookdown.org/yihui/rmarkdown/html-document.html#html-document) or [guide on PDF outputs](https://bookdown.org/yihui/rmarkdown/pdf-document.html#pdf-document).

::: {.callout-important}
    The newly created document contains some example text. Make sure you read it as it gives a good introduction in how to work in Rmarkdown.
:::

#### Knitting

To create an output document you need to render it. The process of rendering is called *knitting* in R terminology. The knitting process takes your settings, code, and text and turns it into a nice document according to your specifications. 

If you have set your output as `html_document` then the knitting process will produce a html file and if you have set it as `pdf_document` it will produce a pdf file. If you do not have a LaTeX distribution installed, you will get an error when trying to knit to a pdf - see @nte-latex-dist.

::: {.callout-note #nte-latex-dist}
# LaTeX distribution for pdf output

In order to create a PDF document from Rmarkdown you will need to have a LaTeX distribution installed on your system. TinyTex is recommended for this, and it can easily be installed from within R using the `tinytex` package:

1. Go to RStudio and run the following code to install the package used for installing TinyTex:

```{r}
#| eval: false
install.packages("tinytex")
```

2. Run the following code to install TinyTex using the `tinytex` package:

```{r}
#| eval: false
tinytex::install_tinytex()
```

3. You should now be able to create a pdf output.

:::

### Mixing code and text in Rmarkdown

