# Multiple linear regression
```{r}
#| label: mlr-setup
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(here)
load(here("data", "Marzipan.RData"))
```

::: {.callout-tip}
## Learning objectives

The learning objectives for this theme are to:

* Graphically show regression problems with several predictors.
* State a statistical model for linear regression with several predictors.
* Compute a regression model.
* Understand the difference between marginal and crude parameter estimates.
* Be able to formulate hypothesis for a given question in relation to regression.

:::

## Reading material

* [Intro to regression ](#sec-intro-to-regression)
* Videos on mutiple linear regression:
   * [Multiple Linear Regression - An Introduction](#sec-mlr-lee-waller)
   * [Multiple Regression, Clearly Explained](#sec-mlr-statquest)
* Chapter 6.1 of [*Introduction to Statistics*](https://02402.compute.dtu.dk/enotes/book-IntroStatistics) by Brockhoff

## Intro to multiple linear regression

Regression models with several predictors are simply extensions of the simple linear regression with a single predictor.

A two-predictor model with $X1$ and $X2$ and a single response variable ($Y$) can be formalized as:

$$
\begin{gather}
Y_i = \alpha + \beta_1 \cdot X1_i + \beta_2 \cdot X2_i + e_i \\
where \ e_i\sim \mathcal{N}(0,\sigma^2) \ and \ independent \\
for \ i=1,...,n
\end{gather}
$$

Where $\alpha$ refers to the level of $Y$ at $X1=X2=0$ and $\beta_1$ and $\beta_2$ are slopes in relation to $X1$ and $X2$.

Each of these three parameters consumes one degree of freedom.

The natural hypothesis is to check whether $X1$ in the presence of $X2$ has an effect on ($Y$), that is: $H0: \beta_1=0$ with the alternative $HA: \beta_1 \neq 0$ (or the other way around). This can be tested with either an F-test or a T-test (yielding the same results).

### Marginal and Crude estimates
In a setup as described above, the investigation of the effect of $X1$ on $Y$ can be done by two approaches:

* A simple regression model: $Y_i = \alpha + \beta X1_i + e_i$
* A multiple regression model: $Y_i = \alpha + \beta_1 X1_i +\beta_2 X2_i + e_i$

Naturally these two approaches yield different results due to the absence/presence of the variable $X2$. From the simple regression model $\beta$ shows the crude effect of $X1$ on $Y$, whereas in the multiple regression model $\beta_1$ shows the marginal effect of $X1$ on $Y$. Sometimes comparison of these models are referred to as *adjusting* for $X2$ or *controlling* for $X2$. These models often elucidate the direct and indirect relation between predictors and responses as is seen in exercise 1.

::: {#exa-nir-of-marcipan-mlr}
<!-- Cross-ref anchor -->
:::
::: {.callout-example}
### Near-infrared spectroscopy of marzipan - regression

In the following example we want to investigate if we can improve a prediction model by using scores from more than one principal component to predict the sugar content in marzipan bread (marcipanbrød).

The PCA model is exactly the same as in @exa-nir-marzipan-least-squares. 

```{r}
#| label: fig-nir-pc1-pc2
#| fig-cap: Score plot of PC1 vs PC2 on a PCA model computed on NIR spectra of marzipan samples
#| code-fold: true

pca <- prcomp(t(X[, -1]), center = T, scale = F)
var_exp <- summary(pca)$importance[2, ] * 100
df <- data.frame(
  pca$x,
  sugar = Y$sugar
)

ggplot(df, aes(PC1, PC2)) + 
  geom_point(size = 2.5, aes(color = sugar)) +
  scale_color_gradient(low = "green", high = "red") +
  labs(
    y = paste("PC2 (", var_exp[2], "%)", sep = ""),
    x = paste("PC1 (", var_exp[1], "%)", sep = ""),
    color = "Sugar content (%)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```


We now make two models:

* A linear regression model on the sugar content versus the scores from PC1 (see @@exa-nir-marzipan-least-squares).
* A linear regression model on the sugar content versus the scores from PC1 and PC2.

```{r}
linreg1 <- lm(sugar ~ PC1, df)
linreg2 <- lm(sugar ~ PC1 + PC2, df)
```

Let us look at the summary of the two models:

```{r}
summary(linreg1)

summary(linreg2)
```

From the summary of `linreg2` we see that the PC2 estimate is slightly significantly different from $0$ ($p-value = 0.01$) and does contribute to improving the model. If we think back on example @exa-nirs-marzipan-pca, we learned that the second principal component had something to do with the color of the marzipan bread. In that regard it does not make a lot of sense that adding a color term (captured by the scores on PC2) into your model improve the prediction of the sugar content. However, when comparing the $R^2$ values for the model including only one component ($R^2 = 0.85$) and a model with two components ($R^2 = 0.88$) it is seen that the improvement does not practically make a difference.

One should always be careful not to build models on meaningless data. PCA is a powerful tool for exploring data, but you need to choose meaningful principal components for your analysis. This is also relevant when selecting variables for the analysis.

Maybe the wavelengths covering the visible part of the spectrum should have been left out of the analysis from the beginning as they do not contain information about the sugar content?

:::
<!-- End example callout -->

## Videos
#### Multiple Linear Regression - An Introduction - *Lee Rusty Waller* {#sec-mlr-lee-waller}
{{< video https://www.youtube.com/watch?v=G4ZlC9zKfII >}}

#### Multiple Regression, Clearly Explained - *StatQuest* {#sec-mlr-statquest}
{{< video https://www.youtube.com/watch?v=EkAQAi3a4js >}}

## Exercises 

::: {#ex-diet-mlr}
<!-- Cross-ref anchor -->
:::
::: {.callout-exercise}
### Diet and fat metabolism - regression with several variables

This exercise examines the relation between a biomarker, dietary intervention and weight in order to disentangle the effects causing elevated levels of cholesterol.

The data for this exercise is the same as for Exercise "Diet and Fat".

In this exercise we are going to focus on predicting cholesterol from both weight and dietary intervention.

::: {.callout-task}
1. Plot, formulate and build univariate models predicting cholesterol level from:
  a. Dietary intervention (`$Fat_Protein`) - A oneway ANOVA model.
  b. Body weight at $14$ weeks (`$bw_w14`) - A regression model.
2. State and test relevant hypothesis for these two models.
3. Comment on the relations between the conclusions of the two models.
4. Make a scatter plot of cholesterol versus weight colored or shaped according to dietary intervention.
5. Make a regression model for cholesterol with two predictors: i) Weight at week $14$ and ii) Dietary intervention, and test the factors.
:::

Note: In contrast to ANOVA with one dependent variable, the order of the dependent variables makes a difference in the `anova()`. In order to surpass this, use the `drop1(aov(),test='F')` function to get results independent of order.

::: {.callout-task}
6. What happens to the significance of the individual factors when you change the analysis as described above?
7. Do you think that the dietary intervention *directly* affects the cholesterol level, or that it is mediated through body weight?
:::

:::
<!-- End exercise callout -->