# Wrangling data in R

```{r}
#| label: programming appendix setup
#| echo: false
#| message: false
#| warning: false


library(dplyr)
library(tidyr)
library(palmerpenguins)

```

## Logical indexing

Logical indexing is one of the most powerful data wrangling techniques. It uses logical operators like `==` (equal to) and `!=` (not equal to) to subset data based on multiply criteria. Below are some examples to explain [indexing](#sec-indexing), [logical operators](#sec-logical-operators), and finally [subsetting data with logical indexing](#sec-logical-indexing).

### Indexing data {#sec-indexing}

If `x` is a vector of the letters A through F in alphabetical order, then we can call the fourth letter (D) by the syntax `x[4]`:

```{r}
x <- c("A", "B", "C", "D", "E", "F")

x[4]

```

If `x` is a matrix (or a dataframe) then we can index the data by the syntax `x[row, column]`. For example

* `x[2, 3]` - Extract the data in row 2, column 3. 
* `x[ , 3]` - Extract all rows in column 3.
* `x[2, ]` - Extract all columns in row 2.

```{r}
# Create a matrix with 3 columns
x <- matrix(c("A", "B", "C", "D", "E", "F"), ncol = 3)

# Print full matrix
x

# Row 2, column 3
x[2, 3]

# All rows, column 3
x[ , 3]

# Row 2, all columns
x[2, ]

```


### Logical operators in R {#sec-logical-operators}

Logical operators are basic statements that return `TRUE` or `FALSE` (@tbl-logical-operators).


| Name                     | Operator |Returns `TRUE`               | Returns `FALSE`            |
|:-------------------------|:---------|:----------------------------|:---------------------------|
| Equal to                 | `==`     | `"A" == "A"`                | `"A" == "B"`               |
| Not equal to             | `!=`     | `"A" != "B"`                | `"A" != "A"`               |
| Less than                | `<`      | `2 < 3"`                    | `3 < 2`                    |
| Greater than             | `>`      | `15 > 10`                   | `10 > 15`                  |
| Less than or equal to    | `<=`     | `10 <= 10` or `9 <= 10`     | `10 <= 9`                  |
| Greater than or equal to | `>=`     | `22 >= 22` or `25 >= 22`    | `20 >= 15`                 |
| In                       | `%in%`   | `"A" %in% c("A", "B" "C")`  | `"F" %in% c("A", "B" "C")` |

: Some of the logical operators available in R {#tbl-logical-operators}

If we have a vector `x` with multiple instances of `A`, `B` and `C` we can index it as:

```{r}
# Create the vector
x <- rep(c("A", "B", "C"), each = 3)

# Print the vector
x

# Check where the elements of x equals "B"
x == "B"
```

If we want to check multiple statements at once we can use the AND `&` and the OR `|` operator. 

```{r}

# Check where the elements of x equals "B" OR "C"
x == "A" | x == "C"

# Check where the elements of x equals "B" AND "C"
x == "A" & x == "C"
```

### Logical indexing of data {#sec-logical-indexing}

It is possible to use a logical vector (i.e. a vector that only contains `TRUE` and `FALSE`) to index data. When passing a logical vector to a data object it returns the data in all the places where the vector equals `TRUE`. 

As an example, we load the `palmerpenguins` dataset, inspect it and do some simple logical indexing.

```{r}

# Inspect the data
head(penguins)

# Extract all rows where species equals "Adelie"
x <- penguins[penguins$species == "Adelie", ]

head(x)

# Extract all rows where species equals "Adelie"
# AND body mass in greater than 4000
x <- penguins[penguins$species == "Adelie" & penguins$body_mass_g > 4000, ]

head(x)

```

## The pipe operator {#sec-pipe-operator}

In R you have the opportunity to use a special syntax called the *pipe operator* `|>`. It can be thought of as a pipe that "sends data downstream" to the next call in your script. 

If we for example want to take the square root of some data `df` and afterwards show the first 6 values of that data with `head()` we can write

```{r}
#| eval: false
df |>
  sqrt() |>
  head()
```

which is equivalent to

```{r}
#| eval: false
head(sqrt(df))
```

This might not make much sense for this example, but when things get a bit more complex the pipe operator can really help by making code easier to read.

If we for example want to format some data `df` and then plot it in ggplot2 we can write the following script

```{r}
#| eval: false

df |>
  pivot_longer(
    cols = everything(),
    names_to = "group",
    names_prefix = "X",
    values_to = "response"
    ) |>
  mutate(group = str_c("Group ", group)) |>
  ggplot(aes(group, response)) +
  geom_boxplot()
```

Without the pipe we would need to create several intermediate variables cluttering up our script and environment in the process. 

## Tidy data

### What is tidy data?

### How to convert to tidy data

## Summarise by group {#sec-summarise-by-group}

Sometimes we want to calculate a statistic per group. There are many different ways of doing this, and in these examples we are going to present a way possible ways.

For these examples we use the *Palmer Penguins* dataset. It consists of some data regarding the bill size, flipper length and body mass of three penguin species across three islands.

Let us start by taking a look at our data. This can be done via the `head()` function or the `glimpse()` function from the Tidyverse.

```{r}
head(penguins) # Show the first 6 rows of dataframe
```

```{r}
glimpse(penguins)
```

### One statistic per group

If we want to summarise one statistic, that could be the mean or the standard deviation, grouped by *species* or *island* (or both), we can do it in the following ways.

::: {.panel-tabset}

#### Using Tidyverse

::: {.callout-important}
If the pipe operator `|>` is unkown to you see @sec-pipe-operator.
:::

To compute the mean of all numeric variables we can use the `summarise()` function from the dplyr package in Tidyverse. Remember to load the tidyverse package (`library(tidyverse)`) - this package includes dplyr as well as a whole lot of other nice packages.

For this example we want the mean for all numeric variables grouped by *species*.

```{r}
penguins |>
  drop_na() |> # Remove rows with missing values
  summarise(
    across(
      .cols = where(is.numeric), # Chose columns that are numeric 
      .fns = mean # Set the function we want to use
           ),
    .by = species # Group by species
    )
```

Be aware that the function drops all non-numeric variables that are not part of the grouping. So the output of the above code is "missing" the *island* and *sex* variables. Also, the `drop_na()` function removes all rows with missing values - if this is not done all columns with missing values will return `NA`.

If we want to group by multiple variables, for example *species* and *island*, we just need to pass a vector of the variables to the `.by = ` argument.

```{r}
penguins |>
  drop_na() |> # Remove rows with missing values
  summarise(
    across(where(is.numeric), mean),
    .by = c(island, species) # Group by island and species
  )
```

#### Using base R

To compute the mean of all variables grouped by *species* we can use the `aggregate()` function. This is included in base R, so there is no need to load any packages. 

```{r}
#| warning: false
# Remove rows with missing values 
penguins_clean <- na.omit(penguins)

aggregate(penguins_clean, list(species = penguins_clean$species), mean)
```

Be aware that all non-numeric variables will return `NA` (and thus a lot of warnings). Also, the `na.omit()` function removes all rows with missing values - if this is not done all columns with missing values will return `NA`.

If we want to group by multiple variables, for example *species* and *island*, we just need to pass another grouping variable to the list.

```{r}
#| warning: false
aggregate(penguins_clean, 
          list(species = penguins_clean$species,
               island = penguins_clean$island), 
          mean)
```

:::

### Create a summary table 

Sometimes we want to create a summary table grouped by some variable. This can be done in the following ways. 

::: {.panel-tabset}

#### Using Tidyverse

::: {.callout-important}
If the pipe operator `|>` is unkown to you see @sec-pipe-operator.
:::

To compute a summary table of a numeric variable we can use the `summarise()` function from the dplyr package in Tidyverse. Remember to load the tidyverse package (`library(tidyverse)`) - this package includes dplyr as well as a whole lot of other nice packages.

For this example we want some summary statistics for the *body_mass_g* variable grouped by *species*.

```{r}
penguins |>
  drop_na() |> # Remove rows with missing values
  summarise(
    N = length(body_mass_g),
    Mean = mean(body_mass_g),
    Median = median(body_mass_g),
    Std = sd(body_mass_g),
    IQR = IQR(body_mass_g),
    .by = species # Group by species
  )
```

Be aware that the `drop_na()` function removes all rows with missing values - if this is not done all columns with missing values will return `NA`.

If we want to group by multiple variables, for example *species* and *island*, we just need to pass another grouping variable to the list.

```{r}
penguins |>
  drop_na() |> # Remove rows with missing values
  summarise(
    N = length(body_mass_g),
    Mean = mean(body_mass_g),
    Median = median(body_mass_g),
    Std = sd(body_mass_g),
    IQR = IQR(body_mass_g),
    .by = c(island, species) # Group by island and species
  )
```

#### Using base R

To compute a summary table of a numeric variable we can use the `aggregate()` function. This is included in base R, so there is no need to load any packages. 

For this example we want some summary statistics for the *body_mass_g* variable grouped by *species*.

```{r}
#| warning: false

# Remove rows with missing values 
penguins_clean <- na.omit(penguins)

# Define variables for grouping
my_groups <- list(species = penguins_clean$species)

# Compute summary stats
penguins_n <- aggregate(penguins_clean, my_groups, length)
penguins_mean <- aggregate(penguins_clean, my_groups, mean)
penguins_median <- aggregate(penguins_clean, my_groups, mean)
penguins_std <- aggregate(penguins_clean, my_groups, mean)
penguins_iqr <- aggregate(penguins_clean, my_groups, mean)

# Collect everything in a dataframe
penguins_summary <- data.frame(
  "Species" = penguins_n$species,
  "N" = penguins_n$body_mass_g,
  "Mean" = penguins_mean$body_mass_g,
  "Median" = penguins_median$body_mass_g,
  "Std" = penguins_std$body_mass_g,
  "IQR" = penguins_iqr$body_mass_g
  )

print(penguins_summary)
```

If we want to group by multiple variables, for example *species* and *island*, we just need to pass another grouping variable to the `my_groups` list.

```{r}
#| warning: false

# Define variables for grouping
my_groups <- list(species = penguins_clean$species,
                  island = penguins_clean$island)

# Compute summary stats
penguins_n <- aggregate(penguins_clean, my_groups, length)
penguins_mean <- aggregate(penguins_clean, my_groups, mean)
penguins_median <- aggregate(penguins_clean, my_groups, mean)
penguins_std <- aggregate(penguins_clean, my_groups, mean)
penguins_iqr <- aggregate(penguins_clean, my_groups, mean)

# Collect everything in a dataframe
penguins_summary <- data.frame(
  "Species" = penguins_n$species,
  "Island" = penguins_n$island,
  "N" = penguins_n$body_mass_g,
  "Mean" = penguins_mean$body_mass_g,
  "Median" = penguins_median$body_mass_g,
  "Std" = penguins_std$body_mass_g,
  "IQR" = penguins_iqr$body_mass_g
)

print(penguins_summary)
```

:::